---
title: "Simulation Challenge"
subtitle: "Generative Models and Monte Carlo Simulation"
format:
  html: default
execute:
  echo: true
  eval: true
---

# 🎲 Simulation Challenge - Monte Carlo Analysis

## Challenge Overview

**Your Mission:** Create a comprehensive Quarto document that simulates one or two investment strategies, analyzes the results, and demonstrates your ability to present counter-intuitive findings compellingly. Then render the document to HTML and deploy it via GitHub Pages from a new repository called "simulationChallenge."

## Investment Game Simulation Analysis

### Original Game Strategy
**Game Rules:** Imagine you are offered the following game and given a $1,000 budget in a special account to play the game: I will flip a coin, and if it comes up heads, we increase your account's balance by 50%; if it comes up tails, we reduce your account's balance by 40%. We are not only doing this once, but we will do it once per year until you turn 55. When you turn 55, you will receive the balance in your account.

## Question 1: Expected Value Analysis

### What is the expected value of your account balance after 1 coin flip?

```{r}
# Set up parameters
initial_balance <- 1000
prob_heads <- 0.5
prob_tails <- 0.5

# Calculate possible outcomes after 1 coin flip
outcome_heads <- initial_balance * 1.5  # +50% if heads
outcome_tails <- initial_balance * 0.6  # -40% if tails

# Calculate expected value
expected_value <- prob_heads * outcome_heads + prob_tails * outcome_tails

cat("=== EXPECTED VALUE CALCULATION ===\n")
cat("Initial balance: $", initial_balance, "\n")
cat("Outcome if heads (+50%): $", outcome_heads, "\n")
cat("Outcome if tails (-40%): $", outcome_tails, "\n")
cat("Probability of heads: ", prob_heads, "\n")
cat("Probability of tails: ", prob_tails, "\n")
cat("\nExpected Value = P(Heads) × Outcome(Heads) + P(Tails) × Outcome(Tails)\n")
cat("Expected Value = ", prob_heads, " × $", outcome_heads, " + ", prob_tails, " × $", outcome_tails, "\n")
cat("Expected Value = $", prob_heads * outcome_heads, " + $", prob_tails * outcome_tails, "\n")
cat("Expected Value = $", expected_value, "\n")
```

### Question 2: Expectation vs Reality

```{r}
# Is the expected value positive or negative?
is_positive <- expected_value > initial_balance
is_negative <- expected_value < initial_balance

cat("\n=== EXPECTATION VS REALITY ===\n")
cat("Expected value: $", expected_value, "\n")
cat("Initial balance: $", initial_balance, "\n")
cat("Expected value is", ifelse(is_positive, "POSITIVE", "NEGATIVE"), "\n")
cat("Expected value is", ifelse(expected_value > initial_balance, "HIGHER", "LOWER"), "than initial balance\n")
cat("Difference from initial: $", round(expected_value - initial_balance, 2), "\n")

# Calculate the expected return percentage
expected_return_pct <- ((expected_value - initial_balance) / initial_balance) * 100
cat("Expected return: ", round(expected_return_pct, 1), "%\n")
```

